# Census Data Wrangling

## Overview

So far, we have used spatial data to characterize access to opioid treatment resources and link those patterns with community-level information. Adding demographic, social, and economic context can help us identify whether disparities in access exist across different populations and better understand factors that may contribute to observed spatial patterns. Many of these community characteristics are available from the U.S. Census Bureau.

The American Community Survey (ACS) is an ongoing survey that provides annually updated estimates of population and housing characteristics. ACS estimates are available as 1-year and 5-year products. We generally recommend using the 5-year estimates for small geographic areas because they provide greater statistical reliability, particularly for less populated areas and smaller population subgroups. In this tutorial, we use the 2020–2024 ACS 5-year estimates, referred to as the 2024 ACS 5-year estimates.

In this tutorial, we demonstrate how to use the tidycensus package to explore and download commonly used ACS variables at multiple geographic levels, both with and without spatial geometries. We will also wrangle these data into formats that can be integrated with other spatial datasets for further analysis. This tutorial focuses on ACS data available through the Census Bureau API. Additional tutorials and documentation for tidycensus are available from the package authors.

Our objectives are thus to:

-   Download ACS data through the Census API
-   Download Census geographic boundaries
-   Explore ACS data at multiple geographic levels
-   Wrangle, clean, and merge Census data for further spatial analysis

## Environment Setup

To replicate the code and functions illustrated in this tutorial, you’ll need to have R and RStudio downloaded and installed on your system. This tutorial assumes some familiarity with the R programming language.

### Input/Output

Our only external input will be a GeoJSON file of the Chicago city boundary. This file can be found here, though all ACS data used in this tutorial will be downloaded directly from the U.S. Census Bureau using the Census API.

-   Chicago City Boundary: [boundaries_chicago.geojson](https://github.com/GeoDaCenter/opioid-environment-toolkit/tree/master/data)

We will generate Census datasets at multiple geographic levels throughout this tutorial. Our final outputs will include:

-   A CSV file and spatial file with selected ACS characteristics at the county level for the state of Illinois.
-   A CSV file and spatial file with selected ACS characteristics for zipcodes within the city of Chicago.

These outputs can be used for further spatial analysis or linked with other datasets using their geographic identifiers.

### Load Libraries

We will use the following packages in this tutorial:

-   `sf`: to manipulate spatial data
-   `tidycensus`: to download Census data using the ACS API
-   `tidyverse`: to manipulate and clean data
-   `tigris`: to download Census TIGER/Line geographic boundaries

First, load the required libraries.

```{r warning = FALSE, message=FALSE}
library(sf)
library(tidycensus)
library(tidyverse)
library(tigris)
```

## Enable Census API Key

To use the Census API, we first need to sign up for an API key. The API key allows the Census Bureau API to identify and authorize requests from your R session. A Census API key can be requested [here](https://api.census.gov/data/key_signup.html).

Once you receive your key, install it by running the code below.

```{r warning = FALSE, message=FALSE}
#census_api_key("yourkeyhere", install = TRUE) # installs the key for future sessions.
```

Setting `install = TRUE` saves the key to your .Renviron file so that it is available for future R sessions.

If you are working on a shared computer or do not want to save your API key, you can use the same function with `install = FALSE`.

To check whether a Census API key is already installed, run:

```{r warning = FALSE, message=FALSE}
Sys.getenv("CENSUS_API_KEY")
```

## Load Data Dynamically

5.4 Load Data Dynamically

We can now start using the tidycensus package to download demographic and socioeconomic data from the U.S. Census Bureau. In this tutorial, we will cover methods to download ACS data at the state, county, census tract, and ZIP Code Tabulation Area (ZCTA) levels. We will also demonstrate how to download the data with and without spatial geometry.

To download a particular variable or table using tidycensus, we need the relevant variable ID. We can identify these IDs by reviewing the variables available through the `load_variables()` function. For details on exploring available ACS variables and identifying their variable IDs, see the Explore Variables section in the Appendix.

We can download ACS variables using the `get_acs()` function. Because ACS data are based on survey samples, estimates are provided along with a margin of error (moe). The `tidycensus` package returns both values for each requested variable in tidy format.

For the examples covered in this tutorial, the four main inputs for the `get_acs()` function are:
-   `geography`: the geographic level at which to retrieve the data (state, county, tract, or zcta)
-   `variables`: a character string or vector of variable IDs to retrieve
-   `year`: the ACS data year to retrieve
-   `geometry`: whether to include spatial geometry in the returned data (TRUE / FALSE)

### State Level

To begin, we can download selected demographic characteristics for all states.

```{r warning = FALSE, message=FALSE}
stateDf <- get_acs(geography = "state",
                   variables = c(totalPop = "B01001_001",
                                 hispanic = "B03003_003",
                                 white = "B02001_002",
                                 black = "B02001_003",
                                 asian = "B02001_005"),
                   year = 2024, geometry = FALSE)

head(stateDf)
```

As we can see, the data are available in tidy format. We can use other tools in the tidyverse to clean and manipulate the data.

```{r warning = FALSE, message=FALSE}
stateDf <- stateDf %>%
           select(GEOID, NAME, variable, estimate) %>%
           pivot_wider(names_from = variable, values_from = estimate) %>%
           mutate(hispanicPr = hispanic / totalPop,
                  whitePr = white / totalPop,
                  blackPr = black / totalPop,
                  asianPr = asian / totalPop) %>%
           select(GEOID, totalPop, hispanicPr, whitePr, blackPr, asianPr)

head(stateDf)
```

### County Level

Similarly, for county-level data:

-   use `geography = "county"` to download all counties in the U.S.
-   use `geography = "county"`, `state = sampleStateName` to download all counties within a state
-   use `geography = "county"`, `state = sampleStateName`, `county = sampleCountyName` to download a specific county

For example, the code below downloads selected demographic characteristics for all counties in Illinois.

```{r warning = FALSE, message=FALSE}
countyDf <- get_acs(geography = "county",
                    variables = c(totalPop = "B01001_001",
                                  hispanic = "B03003_003",
                                  white = "B02001_002",
                                  black = "B02001_003",
                                  asian = "B02001_005"),
                    year = 2024, state = "IL", geometry = FALSE) %>%
            select(GEOID, NAME, variable, estimate) %>%
            pivot_wider(names_from = variable, values_from = estimate) %>%
            mutate(hispanicPr = hispanic / totalPop,
                                whitePr = white / totalPop,
                                blackPr = black / totalPop,
                                asianPr = asian / totalPop) %>%
            select(GEOID, totalPop, hispanicPr, whitePr, blackPr, asianPr)

head(countyDf)
```

### Census tract Level

For census tract-level data, at minimum a state must be provided.

-   use `geography = "tract"`, `state = sampleStateName` to download all census tracts within a state
-   use `geography = "tract"`, `state = sampleStateName`, `county = sampleCountyName` to download all census tracts within a specific county

The code below downloads selected demographic characteristics for all census tracts in Illinois.

```{r warning = FALSE, message=FALSE}
tractDf <- get_acs(geography = "tract",
                   variables = c(totalPop = "B01001_001",
                                 hispanic = "B03003_003",
                                 white = "B02001_002",
                                 black = "B02001_003",
                                 asian = "B02001_005"),
                   year = 2024, state = "IL", geometry = FALSE) %>%
           select(GEOID, NAME, variable, estimate) %>%
           pivot_wider(names_from = variable, values_from = estimate) %>%
           mutate(hispanicPr = hispanic / totalPop,
                  whitePr = white / totalPop,
                  blackPr = black / totalPop,
                  asianPr = asian / totalPop) %>%
           select(GEOID, totalPop, hispanicPr, whitePr, blackPr, asianPr)

head(tractDf)
```

### Zipcode Level

For zipcode-level data, use `geography = "zcta"`. ZIP Code Tabulation Areas (ZCTAs) are Census Bureau approximations of U.S. Postal Service ZIP Codes. Because ZCTAs can cross county and state boundaries, tidycensus retrieves ZCTA data for the entire U.S.

```{r warning = FALSE, message=FALSE}
zctaDf <- get_acs(geography = "zcta",
                  variables = c(totalPop = "B01001_001",
                                hispanic = "B03003_003",
                                white = "B02001_002",
                                black = "B02001_003",
                                asian = "B02001_005"),
                  year = 2024, geometry = FALSE) %>%
          select(GEOID, NAME, variable, estimate) %>%
          pivot_wider(names_from = variable, values_from = estimate) %>%
          mutate(hispanicPr = hispanic / totalPop,
                 whitePr = white / totalPop,
                 blackPr = black / totalPop,
                 asianPr = asian / totalPop) %>%
          select(GEOID, totalPop, hispanicPr, whitePr, blackPr, asianPr)
```

Inspect the data.

```{r warning = FALSE, message=FALSE}
head(zctaDf)
dim(zctaDf)
```

Because ZCTA data are retrieved for the entire nation, we can filter the results to a particular region after downloading them. For example, Chicago ZIP Codes commonly begin with 606, so we can use `str_detect()` as a simple non-spatial filter. If geometry information is available, we can instead overlay the ZCTA geometries with a city boundary to identify the required region. We will illustrate this approach in the Get Geometry section.

In addition to demographic characteristics, ACS data can provide community context that may help us better understand geographic patterns in healthcare access. Here, we use poverty and household vehicle availability as two examples of access-relevant community characteristics.

```{r warning = FALSE, message=FALSE}
zipChicagoDf <- get_acs(geography = "zcta",
                        variables = c(povertyTotal = "B17001_001",
                                      povertyBelow = "B17001_002",
                                      households = "B08201_001",
                                      noVehicle = "B08201_002"),
                        year = 2024, geometry = FALSE) %>%
                select(GEOID, NAME, variable, estimate) %>%
                filter(str_detect(GEOID, "^606")) %>%
                pivot_wider(names_from = variable, values_from = estimate) %>%
                mutate(povertyPr = povertyBelow / povertyTotal,
                       noVehiclePr = noVehicle / households) %>%
                select(GEOID, povertyPr, noVehiclePr)
```

Inspect the Chicago ZCTA data.

```{r warning = FALSE, message=FALSE}
head(zipChicagoDf)
```

### Save Data

We can now save the county and Chicago zipcode datasets as CSV files using the code below.

```{r warning = FALSE, message=FALSE}
write.csv(countyDf, "data/ilcounty_24_demographic.csv")
write.csv(zipChicagoDf, "data/chizips_24_context.csv")
```

## Get Geometry

Geometry and geographic boundaries are key components of American Community Survey data because they provide the spatial framework used to map and analyze Census estimates. Geographic boundaries may be updated over time, and Census data for a given year generally correspond to geographic definitions available for that data vintage. In this tutorial, we use 2024 Census geographic boundaries to correspond with the 2024 ACS 5-year estimates.

The datasets downloaded so far do not include spatial geometry. To conduct spatial analysis, we need to join these data frames to spatially enabled `sf` objects using a common geographic identifier such as GEOID.

We can download Census geographic boundaries using two methods:
-   using `tigris`
-   using `tidycensus`

```{r warning = FALSE, message=FALSE}

```

### Using tigris

To download TIGER/Line geographic boundaries provided by the [U.S. Census Bureau](https://www.census.gov/geographies/mapping-files/time-series/geo/tiger-line-file.html), we can use the `tigris` package. Setting `cb = TRUE` downloads generalized cartographic boundary files, which have less spatial detail and smaller file sizes than full-resolution TIGER/Line files.

For states, counties, and census tracts, we use 2024 boundary files. The 2024 cartographic boundary file for ZCTAs is not currently available, so we use the 2020 ZCTA boundaries.

```{r warning = FALSE, message=FALSE, output=FALSE}
yearToFetch <- 2024

stateShp <- states(year = yearToFetch, cb = TRUE)
countyShp <- counties(year = yearToFetch, state = "IL", cb = TRUE)
tractShp <- tracts(state = "IL", year = yearToFetch, cb = TRUE)
zctaShp <- zctas(year = 2020, cb = TRUE)
```

After downloading the geometry files, we can merge these shapes with the demographic data downloaded in the previous section.

For states:

```{r warning = FALSE, message=FALSE}
# check object types & identifier variable type
# str(stateShp)
# str(stateDf)

stateShp <- merge(stateShp, stateDf, by.x = "STATEFP", by.y = "GEOID", all.x = TRUE)

head(stateShp)
```

Similarly for counties, zctas & census tracts we can use the code below.

```{r warning = FALSE, message=FALSE}
countyShp <- merge(countyShp, countyDf, by.x = "GEOID", by.y = "GEOID", all.x = TRUE)%>%
             select(GEOID, totalPop, hispanicPr, whitePr, blackPr, asianPr)

tractShp <- merge(tractShp, tractDf, by.x = "GEOID", by.y = "GEOID", all.x = TRUE)%>%
            select(GEOID, totalPop, hispanicPr, whitePr, blackPr, asianPr)
```

The 2024 cartographic boundary file for ZCTAs is not currently available, so we use the 2020 ZCTA boundaries, which correspond to the current ZCTA geography used by the 2024 ACS.

```{r warning = FALSE, message=FALSE}
zctaShp <- merge(zctaShp, zctaDf, by.x = "GEOID20", by.y = "GEOID", all.x = TRUE) %>%
              select(GEOID20, totalPop, hispanicPr, whitePr, blackPr, asianPr)
```

### Using tidycensus

The previous method adds an additional step of using the `tigris` package to download geographic boundary files. The `tidycensus` package can also invoke `tigris` within the `get_acs()` function, allowing us to download ACS data with geometry by setting `geometry = TRUE`.

Adding geometry to each variable can increase the size of intermediate datasets and slow down processing, particularly when downloading many variables. For larger requests, we recommend downloading the ACS variables without geometry and then downloading a single variable with `geometry = TRUE`. The resulting spatial object can then be merged with the larger ACS dataset. We illustrate this approach below.

First, download the census tract demographic data without geometry.

```{r warning = FALSE, message=FALSE}
tractDf <- get_acs(geography = "tract",
                   variables = c(totalPop = "B01001_001",
                                 hispanic = "B03003_003",
                                 white = "B02001_002",
                                 black = "B02001_003",
                                 asian = "B02001_005"),
                   year = 2024, state = "IL", geometry = FALSE) %>%
           select(GEOID, NAME, variable, estimate) %>%
           pivot_wider(names_from = variable, values_from = estimate) %>%
           mutate(hispanicPr = hispanic / totalPop,
                  whitePr = white / totalPop,
                  blackPr = black / totalPop,
                  asianPr = asian / totalPop) %>%
           select(GEOID, totalPop, hispanicPr, whitePr, blackPr, asianPr)
```

We can then download a single variable with geometry.

```{r warning = FALSE, message=FALSE}
tractShp <- get_acs(geography = "tract",
                    variables = c(totalPop = "B01001_001"),
                    year = 2024, state = "IL", geometry = TRUE) %>%
            select(GEOID, NAME, variable, estimate) %>%
            pivot_wider(names_from = variable, values_from = estimate)
```

Merge the geometry with the demographic data using GEOID.

```{r warning = FALSE, message=FALSE}
tractShp <- merge(tractShp, tractDf, by.x = "GEOID", by.y = "GEOID", all.x = TRUE)
head(tractShp)
```

To get ZCTA-level data for a specific region, we first download the required dataset for the entire country and then filter the relevant ZCTAs by overlaying the downloaded dataset with the geometry of the region of interest.

Here, we download the poverty and household vehicle availability measures used in the previous section without geometry.

```{r warning = FALSE, message=FALSE}
zctaContextDf <- get_acs(geography = "zcta",
                         variables = c(povertyTotal = "B17001_001",
                                       povertyBelow = "B17001_002",
                                       households = "B08201_001",
                                       noVehicle = "B08201_002"),
                         year = 2024,geometry = FALSE) %>%
               select(GEOID, NAME, variable, estimate) %>%
               pivot_wider(names_from = variable, values_from = estimate) %>%
               mutate(povertyPr = povertyBelow / povertyTotal,
                      noVehiclePr = noVehicle / households) %>%
               select(GEOID, povertyPr, noVehiclePr)
```

Download the 2020 ZCTA boundaries and merge them with the 2024 ACS data.

```{r warning = FALSE, message=FALSE}
zctaShp <- zctas(year = 2020, cb = TRUE)

zctaShp <- merge(zctaShp, zctaContextDf, by.x = "GEOID20", by.y = "GEOID", all.x = TRUE) %>%
           select(GEOID20, povertyPr, noVehiclePr)
```

Inspect the data.

```{r warning = FALSE, message=FALSE}
head(zctaShp)
```

Read in the Chicago city boundary file, which can be download [here](https://github.com/GeoDaCenter/opioid-environment-toolkit/tree/master/data).

```{r warning = FALSE, message=FALSE}
chiCityBoundary <- st_read("data/boundaries_chicago.geojson")
```

Set the same coordinate reference system for both spatial datasets.

```{r warning = FALSE, message=FALSE}
chiCityBoundary <- st_transform(chiCityBoundary, 4326)
zctaShp <- st_transform(zctaShp, 4326)
```

Only keep ZCTAs that intersect the Chicago city boundary.

```{r warning = FALSE, message=FALSE}
zipChicagoShp <- st_intersection(zctaShp, chiCityBoundary)
```

Inspect data.

```{r warning = FALSE, message=FALSE}
head(zipChicagoShp)
```

### Save Data

We previously saved our datasets without geometry information as CSV files. Now we can save the Illinois county demographic data and Chicago ZCTA community context data with geometries as shapefiles using `write_sf()`.

```{r warning = FALSE, message=FALSE}
write_sf(countyShp, "data/ilcounty_24_demographic.shp")
write_sf(zipChicagoShp, "data/chizips_24_context.shp")
```

## Appendix

5.6.1 Explore Variables

Using tidycensus, we can download ACS variables from several types of tables. Some commonly used products include:
-   `Data Profiles`: Broad collections of social, economic, housing, and demographic characteristics, such as Social Characteristics (DP02), Economic Characteristics (DP03), Housing Characteristics (DP04), and Demographic and Housing Estimates (DP05).
-   `Subject Tables`: Topic-specific tables that provide estimates and percentages for subjects such as age and sex (S0101), employment, education, income, and disability.
-   `Detailed Tables`: Detailed ACS tables identified primarily by B and C table IDs. These tables provide the variables used throughout this tutorial.

We can explore the variables available for our year of interest using `load_variables()`. Variable IDs and table structures may change over time, so it is important to check the variables for the ACS year being used.

```{r warning = FALSE, message=FALSE}
sVarNames <- load_variables(2024, "acs5/subject", cache = TRUE)
pVarNames <- load_variables(2024, "acs5/profile", cache = TRUE)
otherVarNames <- load_variables(2024, "acs5", cache = TRUE)

head(pVarNames)
```

A tibble containing table and variable information includes three main columns: name, label, and concept.

name identifies the ACS variable. concept generally identifies the table or topic associated with the variable, while label provides a description of the variable.

We can explore these tibbles to identify the appropriate variable ID for use with `get_acs()`. For example, we can search the Subject Tables for variables related to children under 5 years of age.

```{r warning = FALSE, message=FALSE}
sVarNames %>%
  filter(str_detect(concept, "AGE AND SEX")) %>%
  filter(str_detect(label, "Under 5 years")) %>%
  mutate(label = sub("^Estimate!!", "", label)) %>%
  select(variableId = name, label)
```

We can also search directly within a specific Subject Table, such as S0101.

```{r warning = FALSE, message=FALSE}
sVarNames %>%
  filter(str_sub(name, 1, 5) == "S0101") %>%
  filter(str_detect(label, "Under 5 years")) %>%
  mutate(label = sub("^Estimate!!", "", label)) %>%
  select(variableId = name, label)
```

Similarly, we can explore Data Profile variables. For example, the following code searches for per capita income variables.

```{r warning = FALSE, message=FALSE}
pVarNames %>%
  filter(str_detect(label, "Per capita")) %>%
  mutate(label = sub("^Estimate!!", "", label)) %>%
  select(variable = name, label)
```

We can also search for age-related variables in the Demographic and Housing Estimates profile.

```{r warning = FALSE, message=FALSE}
pVarNames %>%
  filter(str_detect(label, "Under 5 years")) %>%
  mutate(label = sub("^Estimate!!", "", label)) %>%
  select(variable = name, label)
```

Because variable IDs and table structures can change across ACS releases, users should verify the appropriate variable IDs for each year. For analyses across multiple years, Detailed Tables can often provide a more consistent basis for identifying comparable variables. The examples in this tutorial primarily use Detailed Table variables.

For example, we can search the Detailed Tables for per capita income.

```{r warning = FALSE, message=FALSE}
otherVarNames %>%
  filter(str_detect(label, "Per capita income")) %>%
  mutate(label = sub("^Estimate!!", "", label)) %>%
  select(variable = name, label)
```
